---
title: "Modelling of complex, non-linear relationships in time series data while accounting for delayed effects "
subtitle: "Day 2 lab 2: Time series"
author: "Robbie M. Parks"
date: today
format: html
editor: visual
---

## Pre-Class Material

### 0: Preparation

::: cell
```{r}
# 0a Install packages (if needed)
# install.packages("lubridate")

# 0b Load packages
library(readr)
library(dplyr) 
library(lubridate)
library(splines)
library(mgcv)
library(ggplot2)

# 0c Declare directories
library(here)
ProjectFolder <- here("data")
```
:::

### 1: Define Research Question

-   Does daily PM2.5 exposure increase daily cardiovascular hospitalizations in NYC?
-   Examine temporal trends in exposure and outcome before formal modeling.

### 2: Examine NYC Daily Data

::: cell
```{r}
# 2a Read in data
dta <- read_csv(here::here(ProjectFolder, "nyc_time_series.csv"))

# 2b Arrange and inspect
dta <- dta %>% arrange(dateD)
head(dta)
hist(dta$dailyPM)
hist(dta$dailyCVD)
dim(dta)
summary(dta)

# 2c Convert date
dta <- dta %>% mutate(Date = parse_date_time(dateD, "mdy"))

# 2d Day of the week
dta <- dta %>% mutate(DayofWeek = as.character(wday(Date, label = TRUE)))

# 2e Final ordering
dta <- dta %>% arrange(Date)
head(dta)
```
:::

### 3: Plot Time Series of Air Pollution

::: cell
```{r}
# 3a Line plot
ggplot(dta, aes(Date, dailyPM)) + 
  geom_line(color = "darkorange3")
```
:::

::: cell
```{r}
# 3b Smoothed plots
yr_num <- length(unique(year(dta$Date)))
ggplot(dta, aes(Date, dailyPM)) + 
  geom_smooth(color = "darkorange2", method = lm, formula = y ~ ns(x, df = 4 * yr_num)) +
  geom_smooth(color = "green", method = lm, formula = y ~ ns(x, df = 1 * yr_num)) +
  geom_smooth(color = "cyan")
```
:::

### 4: Plot Time Series of Cardiovascular Hospitalizations

::: cell
```{r}
# 4a Points
ggplot(dta, aes(Date, dailyCVD)) + 
  geom_point(color = "mediumpurple3")
```
:::

::: cell
```{r}
# 4b Smoothed
ggplot(dta, aes(Date, dailyCVD)) + 
  geom_smooth(color = "mediumpurple2", method = lm, formula = y ~ ns(x, df = 4 * yr_num)) +
  geom_smooth(color = "green", method = lm, formula = y ~ ns(x, df = 1 * yr_num))
```
:::

### 5: Plot Day of the Week Patterns

::: cell
```{r}
dta$DayofWeek <- factor(dta$DayofWeek, level = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))

dta.dow.cvd <- dta %>% group_by(DayofWeek) %>% summarize(meanCVD = mean(dailyCVD))
dta.dow.pm <- dta %>% group_by(DayofWeek) %>% summarize(meanPM = mean(dailyPM))

ggplot(dta.dow.cvd, aes(DayofWeek, meanCVD)) + geom_bar(stat = "identity")
ggplot(dta.dow.pm, aes(DayofWeek, meanPM)) + geom_bar(stat = "identity")

dta.dow <- bind_rows(
  dta.dow.cvd %>% mutate(VarName = "CVD Hospitalizations") %>% rename(Var = meanCVD),
  dta.dow.pm %>% mutate(VarName = "PM2.5") %>% rename(Var = meanPM)
)

ggplot(dta.dow, aes(DayofWeek, Var)) +
  geom_bar(stat = "identity") +
  facet_grid(VarName ~ ., scales = "free")
```
:::

## Live Session Material

### 6: Define Research Question

-   Estimate association between daily PM2.5 and CVD hospitalizations.
-   Control for temp, RH, seasonal trends, and DOW.

### 7: Time Series Analysis - Lag 0

::: cell
```{r}
mod.ts.p <- glm(dailyCVD ~ dailyPM + ns(dailyTemp, df = 4) +
                  ns(dailyRH, df = 3) + DayofWeek +
                  ns(Date, df = 4 * yr_num),
                family = "poisson", data = dta)

summary(mod.ts.p)

mean(dta$dailyCVD)
var(dta$dailyCVD)

mod.ts.qp <- glm(dailyCVD ~ dailyPM + ns(dailyTemp, df = 4) +
                   ns(dailyRH, df = 3) + DayofWeek +
                   ns(Date, df = 4 * yr_num),
                 family = "quasipoisson", data = dta)

summary(mod.ts.qp)
summary(mod.ts.qp)$dispersion

# RR and CI
coeff.mat <- summary(mod.ts.qp)$coef
coeff_pm <- coeff.mat[2, 1]
se_pm <- coeff.mat[2, 2]

rr <- exp(coeff_pm)
lci <- exp(coeff_pm - 1.96 * se_pm)
uci <- exp(coeff_pm + 1.96 * se_pm)

RR_pm <- paste(round(rr, 4), " (95%CI: ", round(lci, 4), ", ", round(uci, 4), ")", sep = "")
RR_pm

# Percent change for 1-unit and 10-unit
PC_pm.1 <- paste0(round(100 * (exp(coeff_pm) - 1), 2), "% (95%CI: ",
                  round(100 * (exp(coeff_pm - 1.96 * se_pm) - 1), 2), "%, ",
                  round(100 * (exp(coeff_pm + 1.96 * se_pm) - 1), 2), "%)")

PC_pm.10 <- paste0(round(100 * (exp(10 * coeff_pm) - 1), 2), "% (95%CI: ",
                   round(100 * (exp(10 * (coeff_pm - 1.96 * se_pm)) - 1), 2), "%, ",
                   round(100 * (exp(10 * (coeff_pm + 1.96 * se_pm)) - 1), 2), "%)")

PC_pm.1
PC_pm.10
```
:::

### 8: Time Series Analysis - 3-Day Average

::: cell
```{r}
dta <- dta %>% arrange(Date) %>%
  mutate(pm_ave3d = (dailyPM + lag(dailyPM) + lag(dailyPM, 2)) / 3,
         temp_ave3d = (dailyTemp + lag(dailyTemp) + lag(dailyTemp, 2)) / 3,
         rh_ave3d = (dailyRH + lag(dailyRH) + lag(dailyRH, 2)) / 3)

mod.ts.3day.qp <- glm(dailyCVD ~ pm_ave3d +
                        ns(temp_ave3d, df = 4) +
                        ns(rh_ave3d, df = 3) +
                        DayofWeek + ns(Date, df = 4 * yr_num),
                      family = "quasipoisson", data = dta)

summary(mod.ts.3day.qp)

coef.pm <- summary(mod.ts.3day.qp)$coef[2, 1]
se.pm <- summary(mod.ts.3day.qp)$coef[2, 2]

rr <- exp(coef.pm)
lci <- exp(coef.pm - 1.96 * se.pm)
uci <- exp(coef.pm + 1.96 * se.pm)

RR_pm <- paste0(round(rr, 4), " (95% CI: ", round(lci, 4), ", ", round(uci, 4), ")")
RR_pm
```
:::
