---
title: "Introduction to methods in environmental epidemiology"
subtitle: "Day 1 lab 1: Introduction to linear regression in environmental epidemiology"
author: "Robbie M. Parks"
date: today
format: html
editor: visual
---

This lab will involve taking some models and concepts from the Introduction to methods in environmental epidemiology 1 lecture and introduce you to the way linear regression works

::: {.cell}
## 0: Preparation

Load necessary packages
```{r}
library(readr)
library(dplyr)
library(broom)
library(here)
```
:::

::: {.cell}
## 0: Declare Folder Paths
```{r}
ProjectFolder <- here("data")
```
:::

::: {.cell}
## 1: Examine Data
```{r}
dta <- read_csv(here(ProjectFolder, "Lab1_2010_County_Data.csv"))

# Confirm dimensions
print(dim(dta))
print(names(dta))
head(dta)

# View specific variables
print(dta$AvePM[1:10])
print(mode(dta$AvePM))
print(length(dta$AvePM))

# Summary statistics
summary(dta)

# Check specific statistics
print(min(dta$AvePM))
print(max(dta$AvePM))
```
:::

::: {.cell}
## 2: Plotting Data
```{r}
# Histogram
hist(dta$AveBMI, breaks = 20, xlab = "Average BMI", ylab = "Count of Counties", main = "Histogram of Average BMI", las = 1)

# Density
plot(density(dta$AveBMI, na.rm = TRUE), col = "red", las = 1)

# Boxplot
boxplot(split(dta$AvePM, dta$ClimateRegion), xlab = "NOAA Climate Region", ylab = expression("Annual PM"[2.5]))

# Scatterplot
plot(x = dta$AvePM, y = dta$AveBMI, xlab = expression("Average PM"[2.5]), ylab = "Average BMI", main = "Scatterplot: County-average PM vs. BMI")
abline(lm(AveBMI ~ AvePM, data = dta), col = "blue", lwd = 2)
```
:::

## 3: Define research question

In 2010, was county-average BMI associated with county-average annual PM2.5?

::: {.cell}
## 4: Review Data Structure
```{r}
summary(dta$AveBMI)
hist(dta$AveBMI, freq = FALSE, xlab = "Average BMI", ylab = "Density", las = 1)
lines(density(dta$AveBMI, na.rm = TRUE), col = "red")

# Largest BMI
LargestBMI <- dta[dta$AveBMI == max(dta$AveBMI), ]
print(LargestBMI)
```
:::

::: {.cell}
## 5: Fit Regression Model
```{r}
mod <- lm(AveBMI ~ AvePM + PerBlack + PerLatinx + PerAsianAm + MedHInc + 
            MedHVal + LTHS + FemaleUnemp + MaleUnemp + ClimateRegion, 
          data = dta, na.action = na.omit)
```
:::

::: {.cell}
## 6: Review Model Results
```{r}
summary(mod)

# Extract specific components
coeffs <- summary(mod)$coefficients
coeffs[2,]

# Confidence intervals
beta <- coeffs[2,1]
se <- coeffs[2,2]
lci <- beta - 1.96 * se
uci <- beta + 1.96 * se

# Visualize
AvePM.range <- seq(min(dta$AvePM), max(dta$AvePM), length.out = 100)
ChangeBMI <- AvePM.range * beta
plot(AvePM.range, ChangeBMI, xlab = expression("Average PM"[2.5]), ylab = "Associated Change in Average BMI", type = "l", col = "red")
abline(a = 0, b = lci, col = "gray")
abline(a = 0, b = uci, col = "gray")
```
:::

::: {.cell}
## 7: Assess Model Assumptions
```{r}
par(mfrow = c(2, 2))
plot(mod)
```
:::

::: {.cell}
## 8: Examine Effect Modification
```{r}
mod.interaction <- lm(AveBMI ~ AvePM*ClimateRegion + PerBlack + PerLatinx + PerAsianAm + MedHInc + MedHVal + LTHS + FemaleUnemp + MaleUnemp, data = dta, na.action = na.omit)
summary(mod.interaction)

# Compute interaction effects
coef.mat <- summary(mod.interaction)$coefficients
var.mat  <- vcov(mod.interaction)

beta_int <- coef.mat["AvePM", 1] + coef.mat["AvePM:ClimateRegionupper_midwest", 1]
var_int <- var.mat["AvePM", "AvePM"] + var.mat["AvePM:ClimateRegionupper_midwest", "AvePM:ClimateRegionupper_midwest"] + 2 * var.mat["AvePM", "AvePM:ClimateRegionupper_midwest"]
se_int <- sqrt(abs(var_int))
lci_int <- beta_int - 1.96 * se_int
uci_int <- beta_int + 1.96 * se_int

paste("Estimate:", round(beta_int, 3), "95% CI:", round(lci_int, 3), "-", round(uci_int, 3))
```
:::

::: {.cell}
## 9: Breakout Exercises
```{r}
# Example: Unadjusted model
mod.unadj <- lm(AveBMI ~ AvePM, data = dta)
summary(mod.unadj)
```
:::

### Breakout Exercise 1: Change in BMI
> **Question:** What is the expected change in BMI for each 5-unit increase in PM2.5?

> **Answer:** Multiply the PM2.5 coefficient (e.g., `beta`) by 5. If `beta = 0.15`, then 5 * 0.15 = **0.75 unit increase in BMI**.

### Breakout Exercise 2: Explore a Covariate
> **Question:** Fit a model including only AvePM and MedHInc. How does the coefficient for AvePM change?

```{r}
mod.exercise2 <- lm(AveBMI ~ AvePM + MedHInc, data = dta)
summary(mod.exercise2)
```
> **Answer:** Compare the AvePM coefficient to that from the full model. Interpret how adjusting for income alters the estimated effect.

### Breakout Exercise 3: Interpret Assumptions
> **Question:** Which of the model diagnostic plots would indicate potential heteroskedasticity?

> **Answer:** The **Residuals vs Fitted** plot. A funnel shape suggests non-constant variance.

::: {.cell}
## 10: Data Manipulation
```{r}
dta <- read_csv(here(ProjectFolder, "county_pm_bmi_census.csv"))
dta <- dta %>% mutate(AveBMIsq = aveBMI^2, AveObese = if_else(aveBMI >= 30, "average obese", "average not obese"))
```
:::

::: {.cell}
## 11: Summarizing Data
```{r}
MeanPM <- dta %>% summarize(meanPM = mean(avePM.idw, na.rm = TRUE))
MedianPM <- dta %>% summarize(medianPM = median(avePM.idw, na.rm = TRUE))
```
:::

---

This Quarto document covers the introduction, data exploration, regression modeling, assumption checks, and effect modification in the context of environmental epidemiology.