---
title: "Introduction to methods in environmental epidemiology"
subtitle: "Day 1 lab 2: Introduction to quantile regression"
author: "Robbie M. Parks"
date: today
format: html
editor: visual
---


::: {.cell}
## 0: Preparation

Load necessary packages

```{r}
library(readr)
library(dplyr)
library(quantreg)
library(ggplot2)
library(here)
```
:::

::: {.cell}
## 0: Declare Folder Paths
```{r}
ProjectFolder <- here("data")
```
:::

### 1: Load Data

::: {.cell}
```{r}
dta <- read_csv(here(ProjectFolder, "2010_County_Data.csv")) %>%
  filter(complete.cases(.))
```
:::

### 2: Define research question

Does the distribution of county-average BMI vary by annual PM2.5?

### 3: Calculate and Compare Quantiles

::: {.cell}
```{r}
# Identify regions with most observations
table(dta$ClimateRegion)

# Filter for Ohio Valley and Southeast
dta.OhioValley_Southeast <- dta %>% 
  filter(ClimateRegion %in% c("ohio_valley", "southeast"))

# Create separate dataframes for comparison
dta.OhioValley <- filter(dta, ClimateRegion == "ohio_valley")
dta.Southeast <- filter(dta, ClimateRegion == "southeast")

# Calculate quantiles
quantile(dta.OhioValley$AveBMI, c(0.25, 0.5, 0.75))

# Mean vs. outlier sensitivity
OhioValley.Mean <- mean(dta.OhioValley$AveBMI)
dta.OhioValley.noExtreme <- filter(dta.OhioValley, AveBMI < 40)
OhioValley.Mean.noExtreme <- mean(dta.OhioValley.noExtreme$AveBMI)
OhioValley.Mean - OhioValley.Mean.noExtreme

OhioValley.Median <- quantile(dta.OhioValley$AveBMI, 0.5)[[1]]
OhioValley.Median.noExtreme <- quantile(dta.OhioValley.noExtreme$AveBMI, 0.5)[[1]]
OhioValley.Median - OhioValley.Median.noExtreme

# Remove erroneous data
dta <- dta[-c(169),]
rm(list = ls(pattern = "Ohio"))
rm(list = ls(pattern = "dta."))
```
:::


## Live Session Material

### 4: Review Research Question

> Reiteration of question.

### 5: Fit Median Model

::: {.cell}
```{r}
Mod50th <- rq(AveBMI ~ AvePM + PerBlack + PerLatinx + PerAsianAm + 
            MedHInc + MedHVal + LTHS + FemaleUnemp + MaleUnemp + ClimateRegion, 
            data = dta, tau = 0.5)
summary(Mod50th, alpha = 0.05)
Coeff.Mod50th <- summary(Mod50th, alpha = 0.05)$coefficients

# Fit mean model
ModMean <- lm(AveBMI ~ AvePM + PerBlack + PerLatinx + PerAsianAm + 
              MedHInc + MedHVal + LTHS + FemaleUnemp + MaleUnemp + ClimateRegion, 
              data = dta)

# Combine results
coeff.table <- data.frame(Fit.Mod50th = summary(Mod50th)$coefficients[,1],
                           Fit.ModMean = summary(ModMean)$coefficients[,1]) %>%
  mutate(PercentDiff = 100*(Fit.Mod50th - Fit.ModMean)/Fit.ModMean) %>% 
  mutate(across(everything(), round, 3))

# Forest plot
df <- data.frame(Model = c("Mean", "Median"),
                 coeff = c(summary(ModMean)$coefficients[2,1], Coeff.Mod50th[2,1]),
                 lci = c(summary(ModMean)$coefficients[2,1] - 1.96*summary(ModMean)$coefficients[2,2], Coeff.Mod50th[2,2]),
                 uci = c(summary(ModMean)$coefficients[2,1] + 1.96*summary(ModMean)$coefficients[2,2], Coeff.Mod50th[2,3]))

ForestPlotMeanMedian <- ggplot(df, aes(x = Model, y = coeff, ymin = lci, ymax = uci)) + 
  geom_pointrange() + geom_errorbar() + geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Coefficient for PM"[2.5]~" (95% CI)"))
ForestPlotMeanMedian
```
:::

### 6: Compare Association at Different Quantiles

::: {.cell}
```{r}
Mods25.50 <- rq(AveBMI ~ AvePM + PerBlack + PerLatinx + PerAsianAm + 
                  MedHInc + MedHVal + LTHS + FemaleUnemp + MaleUnemp + ClimateRegion, 
                data = dta, tau = c(0.25, 0.5))
summary25.50 <- summary(Mods25.50)

Model25th <- summary25.50[[1]]$coefficients[2,1:3]
Model50th <- summary25.50[[2]]$coefficients[2,1:3]

# Create table and plot
coeff.table <- data.frame(
  ModelName = c("25th", "50th"),
  coeff = c(Model25th[1], Model50th[1]),
  lci = c(Model25th[2], Model50th[2]),
  uci = c(Model25th[3], Model50th[3])
)

ForestPlot.25.50 <- ggplot(coeff.table, aes(x = ModelName, y = coeff, ymin = lci, ymax = uci)) +
  geom_pointrange() + geom_errorbar() + geom_hline(yintercept = 0, lty = 2) +
  xlab("Model Name") + ylab(expression("Coefficient for PM"[2.5]~" (95% CI)"))
ForestPlot.25.50
```
:::

### 6c: Multiple Quantiles

::: {.cell}
```{r}
TauList <- seq(0.1, 0.9, by = 0.1)
qr.Mods <- rq(AveBMI ~ AvePM + PerBlack + PerLatinx + PerAsianAm + 
                 MedHInc + MedHVal + LTHS + FemaleUnemp + MaleUnemp + ClimateRegion, 
               data = dta, tau = TauList)
summary.qr.Mods <- summary(qr.Mods)

coeffs <- sapply(1:9, function(i) summary.qr.Mods[[i]]$coefficients[2,1:3])
coeff.table <- data.frame(t(coeffs))
names(coeff.table) <- c("coeff", "lci", "uci")
coeff.table$ModelName <- paste0(10 * 1:9, "th")

ForestPlot.Mods <- ggplot(coeff.table, aes(x = ModelName, y = coeff, ymin = lci, ymax = uci)) + 
  geom_pointrange() + geom_errorbar() + geom_hline(yintercept = 0, lty = 2) +
  xlab("BMI Quantile") + ylab(expression("Coefficient for PM"[2.5]~" (95% CI)"))
ForestPlot.Mods
```
:::

---

This Quarto document covers the basics of quantile regression with a real world example.
